---
title: "Causal Inference Re-Analysis: National Study of Learning Mindsets"
subtitle: "Carvalho et al. (2019) ACIC 2018 Synthetic Dataset — Portfolio Analysis"
author: "Bhanu Prasad"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "90%"
)

set.seed(42)

# ── Packages ────────────────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(tidyverse)
  library(lme4)
  library(lmerTest)
  library(sandwich)
  library(lmtest)
  library(MatchIt)
  library(cobalt)
  library(glmnet)
  library(survey)
  library(MCMCpack)
  library(coda)
  library(rbounds)
  library(EValue)
  library(kableExtra)
  library(ggdag)
  library(dagitty)
  library(broom)
  library(broom.mixed)
  library(patchwork)
})

# ── Data (single load) ───────────────────────────────────────────────────────
data <- read.csv("data.csv")

# ── Derived variables ────────────────────────────────────────────────────────
# z: keep as numeric 0/1 throughout; coef names will be "z" not "z1"
data$z <- as.integer(data$z)

# race_collapsed: majority = code 4 (~48 % of obs); minority = all others
# Small race categories (< 2 % of 10 391 = < 208 obs) are folded into "other".
data$race_collapsed <- factor(
  ifelse(data$race == 4, "majority", "other"),
  levels = c("other", "majority")
)

# Factor conversions — keep numeric codes, no fabricated labels
data$gender   <- factor(data$gender,  levels = c("1", "2"))
data$fgen     <- factor(data$fgen,    levels = c("0", "1"))
data$urban_f  <- factor(data$urban,   levels = c("0", "1", "2", "3", "4"))
data$schoolid <- factor(data$schoolid)

# ── Shared formula strings ───────────────────────────────────────────────────
cov_rhs    <- "selfrpt + race_collapsed + gender + fgen + urban_f +
               mindset + test + sch_race + pov + size"
ps_form    <- as.formula(paste("z ~", cov_rhs))
out_form   <- as.formula(paste("y ~ z +", cov_rhs))
lmer_form  <- as.formula(paste("y ~ z +", cov_rhs, "+ (1 | schoolid)"))
```

\newpage

# Introduction and Data Context

## Dataset

This analysis applies multiple causal-inference estimators to the **Carvalho, Feller,
Murray, Woody, and Yeager (2019) synthetic dataset** released for the 2018 Atlantic
Causal Inference Conference (ACIC) data competition. The dataset was generated from
the National Study of Learning Mindsets (NSLM), a large-scale randomised trial of
a growth-mindset intervention conducted across 76 U.S. middle schools
(Yeager et al., 2019, *Nature*).

> **Important note on the data generating process.**
> This dataset is *synthetic*. The ACIC organisers simulated outcomes under a known
> causal model using the real NSLM sample frame; the treatment-effect magnitude embedded
> in the simulation (~0.41 SD) does **not** represent the real-world effect.
> Yeager et al. (2019) reported an effect of approximately **0.10 SD on GPA** in the
> actual trial. The purpose of this analysis is methodological: to practice and
> demonstrate causal-inference techniques on a dataset with a known data-generating
> process, not to make empirical claims about the growth-mindset intervention.

## Variables

| Variable | Level | Description |
|----------|-------|-------------|
| `y` | Student | Continuous achievement outcome (standardised) |
| `z` | Student | Treatment indicator (1 = growth-mindset intervention) |
| `selfrpt` | Student | Self-reported prior achievement (1–7 scale) |
| `race` | Student | Race/ethnicity code (15 categories; no public codebook) |
| `gender` | Student | Gender code (1 or 2) |
| `fgen` | Student | First-generation college-student status (0/1) |
| `urban` | Student | School urbanicity (0–4) |
| `mindset` | School | Average school-level mindset score |
| `test` | School | School-level prior test score index |
| `sch_race` | School | School racial composition index |
| `pov` | School | School poverty index |
| `size` | School | School size index |
| `schoolid` | School | School identifier (76 schools) |

: Variable descriptions. School-level variables are constant within school.

## Analytic Approach

Students are **nested within 76 schools**. School-level covariates (`mindset`, `test`,
`sch_race`, `pov`, `size`) are constant within school. Ignoring this clustering
underestimates standard errors. Every estimator below accounts for the clustered
structure either through a multilevel model, cluster-robust sandwich variance, or
a **cluster bootstrap** (resampling schools, not students).

\newpage

# Exploratory Data Analysis

```{r eda-descriptives}
# ── Student-level summary ────────────────────────────────────────────────────
desc_tbl <- data %>%
  group_by(z) %>%
  summarise(
    N          = n(),
    Mean_y     = round(mean(y), 3),
    SD_y       = round(sd(y), 3),
    Mean_selfrpt = round(mean(selfrpt), 2),
    Pct_fgen   = round(100 * mean(fgen == "1"), 1),
    Pct_majority_race = round(100 * mean(race_collapsed == "majority"), 1),
    .groups = "drop"
  ) %>%
  mutate(z = ifelse(z == 1, "Treatment", "Control"))

kable(desc_tbl, booktabs = TRUE, col.names = c("Group", "N", "Mean y",
      "SD y", "Mean selfrpt", "% First-gen", "% Race-majority"),
      caption = "Descriptive statistics by treatment group.") %>%
  kable_styling(latex_options = c("hold_position"))
```

```{r eda-race-table}
# Race code distribution — keep numeric, flag small categories
race_tbl <- data %>%
  count(race) %>%
  mutate(pct = round(100 * n / nrow(data), 1),
         note = ifelse(pct < 2, "< 2 % — folded into 'other'", "")) %>%
  arrange(desc(n))

kable(race_tbl, booktabs = TRUE,
      col.names = c("Race code", "N", "Percent", "Note"),
      caption = "Race code distribution. No public codebook exists for these codes.
                 Race code 4 (48\\%) is treated as the majority category; all others
                 collapse to 'other' for subgroup analysis.") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

```{r eda-plots, fig.height=4}
p1 <- ggplot(data, aes(x = y, fill = factor(z, labels = c("Control","Treatment")))) +
  geom_density(alpha = 0.55, colour = NA) +
  scale_fill_manual(values = c("#4477AA", "#EE6677"), name = NULL) +
  labs(title = "Outcome distribution by treatment",
       x = "Achievement (y)", y = "Density") +
  theme_bw(base_size = 11) +
  theme(legend.position = "top")

p2 <- ggplot(data, aes(x = factor(z, labels = c("Control","Treatment")),
                        y = selfrpt, fill = factor(z))) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  scale_fill_manual(values = c("#4477AA", "#EE6677"), guide = "none") +
  labs(title = "Self-reported prior achievement",
       x = NULL, y = "selfrpt (1-7)") +
  theme_bw(base_size = 11)

p1 + p2
```

```{r eda-school-level, fig.height=3.5}
# Schools: treatment take-up rate varies
school_tbl <- data %>%
  group_by(schoolid) %>%
  summarise(treat_rate = mean(z), n = n(), .groups = "drop")

ggplot(school_tbl, aes(x = treat_rate)) +
  geom_histogram(bins = 20, fill = "#4477AA", colour = "white") +
  labs(title = "Treatment take-up rate across 76 schools",
       x = "Proportion treated in school", y = "Number of schools") +
  theme_bw(base_size = 11)
```

\newpage

# Causal Framework

## Identification Assumptions

1. **Ignorability (conditional unconfoundedness):** $Y(0), Y(1) \perp Z \mid X$.
   All relevant confounders of treatment assignment are captured in the observed
   covariates. Assessed via propensity-score overlap and sensitivity analysis.

2. **Positivity (overlap):** $0 < P(Z=1 \mid X) < 1$ for all covariate values in
   the support. Verified via propensity-score distributions.

3. **SUTVA:** No interference between students and a single well-defined version of
   treatment. Plausible within-school, though school-level spillovers cannot be ruled out.

4. **Partial interference / clustering:** Students within the same school share the
   school-level intervention environment. Treatment assignment is at the school level
   for some components. All SEs account for this via cluster-robust or multilevel methods.

## Directed Acyclic Graph

```{r dag, fig.height=5.5}
dag <- dagify(
  y       ~ z + selfrpt + race + gender + fgen + mindset + test + sch_race + pov + size,
  z       ~ selfrpt + race + gender + fgen + urban + mindset + test + sch_race + pov + size,
  selfrpt ~ race + gender + fgen + mindset,
  labels = c(
    y       = "Achievement (y)",
    z       = "Mindset\nIntervention",
    selfrpt = "Self-reported\nExpectations",
    race    = "Race code",
    gender  = "Gender",
    fgen    = "First-gen\nStatus",
    mindset = "School\nMindset",
    test    = "School\nTest Score",
    sch_race= "School\nRacial Comp.",
    pov     = "School\nPoverty",
    size    = "School\nSize",
    urban   = "Urbanicity"
  ),
  outcome = "y",
  exposure = "z"
)

ggdag(dag, use_labels = "label", text = FALSE) +
  theme_dag() +
  labs(title = "Assumed causal structure for the growth-mindset intervention")
```

\newpage

# Propensity Score Estimation and Weight Diagnostics

```{r ps-model}
# ── Logistic PS model ────────────────────────────────────────────────────────
ps_model <- glm(ps_form, family = binomial(), data = data)
data$pscore <- predict(ps_model, type = "response")

# ── Stabilised ATE weights ───────────────────────────────────────────────────
# Stabilised weights reduce variance vs. raw Horvitz–Thompson weights.
p_z <- mean(data$z)
data$weight_stab <- ifelse(
  data$z == 1,
  p_z / data$pscore,
  (1 - p_z) / (1 - data$pscore)
)

# ── Weight diagnostics ───────────────────────────────────────────────────────
w <- data$weight_stab
ess <- sum(w)^2 / sum(w^2)

w_summary <- data.frame(
  Stat  = c("Min", "Q1", "Median", "Mean", "Q3", "Max",
            "Max / Mean ratio", "Effective Sample Size (ESS)",
            "ESS / N (%)"),
  Value = c(round(min(w), 3), round(quantile(w, 0.25), 3),
            round(median(w), 3), round(mean(w), 3),
            round(quantile(w, 0.75), 3), round(max(w), 3),
            round(max(w) / mean(w), 1),
            round(ess, 0),
            round(100 * ess / nrow(data), 1))
)

kable(w_summary, booktabs = TRUE,
      caption = "Stabilised ATE weight diagnostics. Max/Mean $< 10$ indicates
                 no extreme weights requiring trimming.") %>%
  kable_styling(latex_options = "hold_position")
```

```{r ps-trim-check}
# Trim weights only if max > 10x mean (common heuristic)
max_mean_ratio <- max(w) / mean(w)
if (max_mean_ratio > 10) {
  trim_val <- quantile(w, 0.99)
  data$weight_stab <- pmin(data$weight_stab, trim_val)
  message("Weights trimmed at 99th percentile (", round(trim_val, 3), ")")
} else {
  message("No trimming required: max/mean ratio = ", round(max_mean_ratio, 2))
}
```

```{r ps-plots, fig.height=4}
p_ps1 <- ggplot(data, aes(x = pscore,
                            fill = factor(z, labels = c("Control","Treatment")))) +
  geom_density(alpha = 0.55, colour = NA) +
  scale_fill_manual(values = c("#4477AA", "#EE6677"), name = NULL) +
  labs(title = "Propensity score overlap",
       x = "Estimated P(Z=1 | X)", y = "Density") +
  theme_bw(base_size = 11) + theme(legend.position = "top")

p_ps2 <- ggplot(data, aes(x = weight_stab,
                            fill = factor(z, labels = c("Control","Treatment")))) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity", colour = NA) +
  scale_fill_manual(values = c("#4477AA", "#EE6677"), name = NULL) +
  labs(title = "Stabilised weight distribution",
       x = "ATE weight", y = "Count") +
  theme_bw(base_size = 11) + theme(legend.position = "top")

p_ps1 + p_ps2
```

```{r love-plot, fig.height=5}
# Love plot: balance before and after IPW
bal_unweighted <- bal.tab(ps_form, data = data, estimand = "ATE",
                           m.threshold = 0.1)
love.plot(bal_unweighted, stat = "mean.diffs", threshold = 0.1,
          abs = TRUE, var.order = "unadjusted",
          title = "Standardised mean differences: unweighted vs. IPW-weighted",
          sample.names = c("Unweighted", "Weighted"),
          colors = c("#888888", "#EE6677")) +
  theme_bw(base_size = 10)
```

\newpage

# Average Treatment Effect Estimation

All models include the full covariate set: `selfrpt`, `race_collapsed`, `gender`,
`fgen`, `urban_f` (student-level) and `mindset`, `test`, `sch_race`, `pov`, `size`
(school-level).

## Multilevel Outcome Model (Primary Estimator)

```{r ate-lmer}
lmer_fit <- lmer(lmer_form, data = data, REML = FALSE)
lmer_summ <- summary(lmer_fit)

lmer_coef  <- fixef(lmer_fit)["z"]
lmer_se    <- lmer_summ$coefficients["z", "Std. Error"]
lmer_df    <- lmer_summ$coefficients["z", "df"]
lmer_ci    <- confint(lmer_fit, parm = "z", method = "Wald")

cat(sprintf(
  "lmer ATE = %.4f  SE = %.4f  95%% CI [%.4f, %.4f]  df = %.1f\n",
  lmer_coef, lmer_se, lmer_ci[1], lmer_ci[2], lmer_df
))

# Random-effect variance
vc <- as.data.frame(VarCorr(lmer_fit))
icc <- vc$vcov[1] / (vc$vcov[1] + vc$vcov[2])
cat(sprintf("School-level ICC = %.3f\n", icc))
```

The intraclass correlation (ICC) quantifies how much of the residual outcome variance
is between schools. An ICC around 0.10–0.20 is common in educational data and confirms
that ignoring school clustering would have meaningfully underestimated standard errors.

## OLS Regression Adjustment

```{r ate-ols}
ols_fit <- lm(out_form, data = data)

# Naive SE
ols_naive_ci  <- confint(ols_fit)["z", ]
ols_est       <- coef(ols_fit)["z"]

# Cluster-robust SE via sandwich::vcovCL
vcov_cl   <- vcovCL(ols_fit, cluster = ~schoolid)
ols_cl_ci <- coefci(ols_fit, parm = "z", vcov. = vcov_cl)

cat(sprintf("OLS ATE = %.4f\n  Naive    95%% CI [%.4f, %.4f]\n  Cluster  95%% CI [%.4f, %.4f]\n",
    ols_est, ols_naive_ci[1], ols_naive_ci[2],
    ols_cl_ci[1], ols_cl_ci[2]))
```

## IPW with Cluster-Robust SE

```{r ate-ipw}
# Naive IPW (ids = ~1, no clustering)
design_naive <- svydesign(ids = ~1, data = data, weights = ~weight_stab)
ipw_naive    <- svyglm(y ~ z, design = design_naive)
ipw_est      <- coef(ipw_naive)["z"]
ipw_naive_ci <- confint(ipw_naive)["z", ]

# Cluster-robust IPW (ids = ~schoolid)
design_cl <- svydesign(ids = ~schoolid, data = data, weights = ~weight_stab)
ipw_cl    <- svyglm(y ~ z, design = design_cl)
ipw_cl_ci <- confint(ipw_cl)["z", ]

cat(sprintf("IPW ATE = %.4f\n  Naive   95%% CI [%.4f, %.4f]\n  Cluster 95%% CI [%.4f, %.4f]\n",
    ipw_est, ipw_naive_ci[1], ipw_naive_ci[2],
    ipw_cl_ci[1], ipw_cl_ci[2]))
```

## Doubly Robust Estimation

```{r ate-dr}
# DR: outcome model with covariates, inside cluster-robust IPW design
dr_cl    <- svyglm(as.formula(paste("y ~ z +", cov_rhs)), design = design_cl)
dr_est   <- coef(dr_cl)["z"]
dr_cl_ci <- confint(dr_cl)["z", ]

# Naive version (ids = ~1)
dr_naive    <- svyglm(as.formula(paste("y ~ z +", cov_rhs)), design = design_naive)
dr_naive_ci <- confint(dr_naive)["z", ]

cat(sprintf("DR ATE = %.4f\n  Naive   95%% CI [%.4f, %.4f]\n  Cluster 95%% CI [%.4f, %.4f]\n",
    dr_est, dr_naive_ci[1], dr_naive_ci[2],
    dr_cl_ci[1], dr_cl_ci[2]))
```

## Lasso-IPW

```{r ate-lasso}
# Build model matrix for Lasso propensity model
X_ps <- model.matrix(ps_form, data = data)[, -1]
y_z  <- data$z

set.seed(42)
cv_lasso  <- cv.glmnet(X_ps, y_z, family = "binomial", alpha = 1, nfolds = 10)
data$pscore_lasso <- as.vector(predict(cv_lasso, newx = X_ps,
                                        type = "response", s = "lambda.min"))

data$weight_lasso <- ifelse(
  data$z == 1, p_z / data$pscore_lasso,
  (1 - p_z) / (1 - data$pscore_lasso)
)

design_lasso  <- svydesign(ids = ~schoolid, data = data, weights = ~weight_lasso)
ipw_lasso_fit <- svyglm(y ~ z, design = design_lasso)
lasso_est     <- coef(ipw_lasso_fit)["z"]
lasso_cl_ci   <- confint(ipw_lasso_fit)["z", ]

# Naive version for comparison
design_lasso_naive <- svydesign(ids = ~1, data = data, weights = ~weight_lasso)
ipw_lasso_naive    <- svyglm(y ~ z, design = design_lasso_naive)
lasso_naive_ci     <- confint(ipw_lasso_naive)["z", ]

cat(sprintf("Lasso-IPW ATE = %.4f\n  Naive   95%% CI [%.4f, %.4f]\n  Cluster 95%% CI [%.4f, %.4f]\n",
    lasso_est, lasso_naive_ci[1], lasso_naive_ci[2],
    lasso_cl_ci[1], lasso_cl_ci[2]))
```

## Propensity Score Matching + Regression Adjustment

```{r ate-psm}
# 1:1 nearest-neighbour matching on logistic PS
match_fit <- matchit(ps_form, data = data, method = "nearest",
                     ratio = 1, replace = FALSE)
matched   <- match.data(match_fit)

# Regression adjustment on matched sample with cluster-robust SE
psm_reg   <- lm(out_form, data = matched, weights = weights)
psm_est   <- coef(psm_reg)["z"]

vcov_psm_naive <- vcov(psm_reg)
vcov_psm_cl    <- vcovCL(psm_reg, cluster = ~schoolid)

psm_naive_ci <- coefci(psm_reg, parm = "z")["z", ]
psm_cl_ci    <- coefci(psm_reg, parm = "z", vcov. = vcov_psm_cl)["z", ]

cat(sprintf("PSM+Reg ATE = %.4f\n  Naive   95%% CI [%.4f, %.4f]\n  Cluster 95%% CI [%.4f, %.4f]\n",
    psm_est, psm_naive_ci[1], psm_naive_ci[2],
    psm_cl_ci[1], psm_cl_ci[2]))

# Balance after matching
bal_after <- bal.tab(match_fit, m.threshold = 0.1)
love.plot(bal_after, stat = "mean.diffs", threshold = 0.1, abs = TRUE,
          title = "Balance before and after nearest-neighbour matching",
          sample.names = c("Unmatched", "Matched"),
          colors = c("#888888", "#4477AA")) +
  theme_bw(base_size = 10)
```

\newpage

# Cluster-Robust Inference: Naive vs. Cluster-Robust CIs

## Cluster Bootstrap

```{r cluster-bootstrap}
# ── Helper: build one bootstrap dataset (resample schools) ───────────────────
draw_cluster_boot <- function(dat) {
  schools   <- levels(dat$schoolid)
  n_schools <- length(schools)
  sampled   <- sample(schools, n_schools, replace = TRUE)
  boot_list <- lapply(seq_along(sampled), function(j) {
    d <- dat[dat$schoolid == sampled[j], ]
    d$schoolid <- factor(j)   # unique ID for the replicated school
    d
  })
  do.call(rbind, boot_list)
}

# ── Estimator functions ──────────────────────────────────────────────────────
est_ols <- function(d) {
  coef(lm(out_form, data = d))["z"]
}

est_ipw <- function(d) {
  ps   <- glm(ps_form, family = binomial(), data = d)
  d$ps <- predict(ps, type = "response")
  pz   <- mean(d$z)
  d$w  <- ifelse(d$z == 1, pz / d$ps, (1 - pz) / (1 - d$ps))
  coef(lm(y ~ z, data = d, weights = d$w))["z"]
}

est_dr <- function(d) {
  ps   <- glm(ps_form, family = binomial(), data = d)
  d$ps <- predict(ps, type = "response")
  pz   <- mean(d$z)
  d$w  <- ifelse(d$z == 1, pz / d$ps, (1 - pz) / (1 - d$ps))
  coef(lm(out_form, data = d, weights = d$w))["z"]
}

est_lmer <- function(d) {
  # suppressWarnings to suppress convergence messages in replicates
  m <- suppressWarnings(
    lmer(lmer_form, data = d, REML = FALSE,
         control = lmerControl(optimizer = "bobyqa"))
  )
  fixef(m)["z"]
}

# ── Run cluster bootstrap (500 reps OLS/IPW/DR; 200 reps lmer) ──────────────
set.seed(42)
N_BOOT_FAST <- 500
N_BOOT_SLOW <- 200

run_boot <- function(estimator_fn, n_reps, dat) {
  vapply(seq_len(n_reps), function(i) {
    bd <- draw_cluster_boot(dat)
    tryCatch(estimator_fn(bd), error = function(e) NA_real_)
  }, numeric(1))
}

boot_ols  <- run_boot(est_ols,  N_BOOT_FAST, data)
boot_ipw  <- run_boot(est_ipw,  N_BOOT_FAST, data)
boot_dr   <- run_boot(est_dr,   N_BOOT_FAST, data)
boot_lmer <- run_boot(est_lmer, N_BOOT_SLOW, data)

ci_boot <- function(x, probs = c(0.025, 0.975)) {
  quantile(x, probs, na.rm = TRUE)
}
```

## Comparison Table: Naive vs. Cluster-Robust 95% CIs

```{r ci-comparison-table}
make_ci_str <- function(lo, hi) sprintf("[%.3f, %.3f]", lo, hi)

ci_table <- tribble(
  ~Method, ~Estimate,
  ~Naive_CI, ~ClusterRobust_CI,

  "OLS Regression",
  round(ols_est, 3),
  make_ci_str(ols_naive_ci[1], ols_naive_ci[2]),
  make_ci_str(ci_boot(boot_ols)[1], ci_boot(boot_ols)[2]),

  "IPW (stabilised)",
  round(ipw_est, 3),
  make_ci_str(ipw_naive_ci[1], ipw_naive_ci[2]),
  make_ci_str(ci_boot(boot_ipw)[1], ci_boot(boot_ipw)[2]),

  "Doubly Robust",
  round(dr_est, 3),
  make_ci_str(dr_naive_ci[1], dr_naive_ci[2]),
  make_ci_str(ci_boot(boot_dr)[1], ci_boot(boot_dr)[2]),

  "Lasso-IPW",
  round(lasso_est, 3),
  make_ci_str(lasso_naive_ci[1], lasso_naive_ci[2]),
  make_ci_str(lasso_cl_ci[1], lasso_cl_ci[2]),

  "PSM + Regression",
  round(psm_est, 3),
  make_ci_str(psm_naive_ci[1], psm_naive_ci[2]),
  make_ci_str(psm_cl_ci[1], psm_cl_ci[2]),

  "Multilevel (lmer)",
  round(lmer_coef, 3),
  "— (inherently multilevel)",
  make_ci_str(ci_boot(boot_lmer)[1], ci_boot(boot_lmer)[2])
)

kable(ci_table, booktabs = TRUE,
      col.names = c("Method", "Estimate", "Naive 95% CI", "Cluster-Robust 95% CI"),
      caption = "ATE estimates with naive (iid) and cluster-robust 95\\% CIs.
                 Naive CIs treat all 10{,}391 students as independent; cluster-robust
                 CIs account for the 76-school multilevel structure via cluster
                 bootstrap (500 reps OLS/IPW/DR; 200 reps lmer) or sandwich SE.
                 All estimates are in SD units of the standardised outcome.") %>%
  kable_styling(latex_options = c("hold_position", "scale_down")) %>%
  column_spec(4, bold = TRUE)
```

```{r naive-vs-cluster-widths, fig.height=3.5}
# Visualise how much wider cluster-robust CIs are
width_df <- tribble(
  ~Method, ~Naive_Width, ~Cluster_Width,
  "OLS",      diff(ols_naive_ci),         diff(ci_boot(boot_ols)),
  "IPW",      diff(ipw_naive_ci),         diff(ci_boot(boot_ipw)),
  "DR",       diff(dr_naive_ci),          diff(ci_boot(boot_dr)),
  "lmer",     diff(as.numeric(lmer_ci)),  diff(ci_boot(boot_lmer))
) %>%
  pivot_longer(cols = c(Naive_Width, Cluster_Width),
               names_to = "Type", values_to = "Width") %>%
  mutate(Type = recode(Type,
                        Naive_Width   = "Naive (iid)",
                        Cluster_Width = "Cluster-robust"))

ggplot(width_df, aes(x = Method, y = Width, fill = Type)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("#4477AA", "#EE6677"), name = NULL) +
  labs(title = "CI width: naive vs. cluster-robust",
       subtitle = "Cluster-robust CIs are wider because students within a school share correlated errors",
       x = NULL, y = "95% CI width (SD units)") +
  theme_bw(base_size = 11) + theme(legend.position = "top")
```

\newpage

# Bayesian Analysis

```{r bayesian-fit, cache=FALSE}
# MCMCregress with non-informative priors
# burnin=1000, mcmc=10000, thin=10 → 1000 effective draws
set.seed(42)
bayes_fit <- MCMCregress(
  formula = out_form,
  data    = data,
  burnin  = 1000,
  mcmc    = 10000,
  thin    = 10,
  seed    = 42
)

# The treatment coefficient is named "z" (numeric 0/1, not factor)
z_col <- grep("^z$", colnames(bayes_fit), value = TRUE)
if (length(z_col) == 0) {
  # Fallback: pick the column whose name starts with "z"
  z_col <- grep("^z", colnames(bayes_fit), value = TRUE)[1]
}

ate_posterior <- as.numeric(bayes_fit[, z_col])
post_ci       <- quantile(ate_posterior, c(0.025, 0.5, 0.975))
post_mean     <- mean(ate_posterior)
post_sd       <- sd(ate_posterior)

cat(sprintf(
  "Bayesian ATE\n  Posterior mean = %.4f\n  Posterior SD   = %.4f\n  95%% CrI       [%.4f, %.4f]\n",
  post_mean, post_sd, post_ci[1], post_ci[3]
))
```

```{r bayesian-density-plot, fig.height=4}
ggplot(data.frame(ATE = ate_posterior), aes(x = ATE)) +
  geom_density(fill = "#4477AA", alpha = 0.6, colour = "#22355A") +
  geom_vline(xintercept = post_ci[c(1, 3)], linetype = "dashed",
             colour = "#EE6677", linewidth = 0.8) +
  geom_vline(xintercept = post_ci[2], linetype = "solid",
             colour = "#AA3333", linewidth = 1) +
  annotate("text", x = post_ci[2] + 0.003, y = Inf, vjust = 1.5, hjust = 0,
           label = sprintf("Median = %.3f", post_ci[2]), size = 3.5) +
  labs(title = "Posterior distribution of ATE (Bayesian linear regression)",
       subtitle = "Dashed lines: 95% credible interval; solid line: posterior median",
       x = "ATE (SD units)", y = "Posterior density") +
  theme_bw(base_size = 11)
```

Quick MCMC diagnostics (full trace plots and Geweke Z-scores in Appendix A):

```{r bayesian-quick-diag}
# Effective sample size and Geweke diagnostic for z coefficient
ess_z    <- effectiveSize(bayes_fit[, z_col])
geweke_z <- geweke.diag(bayes_fit[, z_col])$z

cat(sprintf(
  "z coefficient: ESS = %.0f (of 1000 draws)   Geweke Z = %.3f %s\n",
  ess_z,
  geweke_z,
  ifelse(abs(geweke_z) < 1.96, "(no convergence concern)", "(FLAG: |Z| > 1.96)")
))
```

\newpage

# Sensitivity Analysis

## Rosenbaum Bounds

Rosenbaum bounds ask: how large would unmeasured confounding have to be—in terms
of the odds-ratio advantage Γ that a confounder provides for treatment assignment—
before the observed effect becomes non-significant?

```{r rosenbaum-bounds, fig.height=4}
# Extract matched pairs from the 1:1 PSM match (subclass = pair index)
pairs_df <- matched %>%
  group_by(subclass) %>%
  filter(n() == 2 && length(unique(z)) == 2) %>%  # complete pairs only
  summarise(
    y_treat = y[z == 1],
    y_ctrl  = y[z == 0],
    .groups = "drop"
  )

# Run Rosenbaum bounds from Gamma = 1.0 to 3.0 in steps of 0.1
# psens returns a table of upper/lower bound p-values for each Gamma level
sens_out <- psens(pairs_df$y_treat, pairs_df$y_ctrl,
                  Gamma = 3.0, GammaInc = 0.1)

bounds_df <- as.data.frame(sens_out$bounds)
names(bounds_df) <- c("Gamma", "p_lower", "p_upper")

# Breakpoint: first Gamma where upper-bound p > 0.05
break_idx <- which(bounds_df$p_upper > 0.05)[1]
if (is.na(break_idx)) {
  gamma_break <- "> 3.0"
  cat("Effect remains significant at Gamma = 3.0 (p_upper =",
      round(tail(bounds_df$p_upper, 1), 4), ")\n")
} else {
  gamma_break <- bounds_df$Gamma[break_idx]
  cat(sprintf("Significance breaks at Gamma = %.1f (p_upper = %.4f)\n",
              gamma_break, bounds_df$p_upper[break_idx]))
}

# Plot
ggplot(bounds_df, aes(x = Gamma)) +
  geom_line(aes(y = p_upper, colour = "Upper bound"), linewidth = 1) +
  geom_line(aes(y = p_lower, colour = "Lower bound"), linewidth = 1,
            linetype = "dashed") +
  geom_hline(yintercept = 0.05, linetype = "dotted", colour = "grey40") +
  annotate("text", x = 1.1, y = 0.06, label = "p = 0.05", size = 3,
           colour = "grey40") +
  scale_colour_manual(values = c("Upper bound" = "#EE6677",
                                  "Lower bound" = "#4477AA"),
                       name = NULL) +
  scale_x_continuous(breaks = seq(1, 3, 0.25)) +
  labs(title = "Rosenbaum bounds sensitivity analysis",
       subtitle = sprintf("Significance breaks at Gamma = %s", gamma_break),
       x = expression(Gamma ~ "(unmeasured confounding odds ratio)"),
       y = "Wilcoxon signed-rank p-value (bound)") +
  theme_bw(base_size = 11) + theme(legend.position = "top")
```

**Interpretation.** The result is robust to unmeasured confounding up to
Γ = `r gamma_break`. An unmeasured confounder would have to multiply the
odds of treatment assignment by at least `r gamma_break` to explain away the
observed effect — a very strong confounder.

## E-value

```{r e-value}
# E-value: minimum confounding strength (as a risk-ratio) needed to explain
# away the effect. Uses the lmer estimate as the primary ATE.
sd_y <- sd(data$y)

evalue_result <- evalues.OLS(
  est   = lmer_coef,
  se    = lmer_se,
  sd    = sd_y,
  delta = 1,
  true  = 0
)

print(evalue_result)

ev_point <- evalue_result["E-values", "point"]
ev_lower <- evalue_result["E-values", "lower"]

cat(sprintf(
  "\nE-value (point estimate): %.2f\nE-value (lower CI):       %.2f\n",
  ev_point, ev_lower
))
```

**Interpretation.** An unmeasured confounder would need to be associated with
treatment assignment by a risk-ratio of at least **`r round(ev_point, 2)`** and
with the outcome by the same factor to fully explain away the observed ATE.
Even the lower confidence limit of the E-value is **`r round(ev_lower, 2)`**,
indicating substantial robustness.

## Model Specification Robustness

```{r sensitivity-models}
# Minimal model (student-level covariates only, no multilevel structure)
mod_minimal <- lm(y ~ z + selfrpt + gender + fgen, data = data)
est_minimal <- coef(mod_minimal)["z"]
ci_minimal  <- coefci(mod_minimal, parm = "z",
                       vcov. = vcovCL(mod_minimal, cluster = ~schoolid))

# Full multilevel model (Section 5 primary)
# lmer_coef already computed

spec_df <- data.frame(
  Model    = c("Minimal (student covariates only)", "Full multilevel (lmer)"),
  Estimate = round(c(est_minimal, lmer_coef), 3),
  CI_lo    = round(c(ci_minimal[1], lmer_ci[1]), 3),
  CI_hi    = round(c(ci_minimal[2], lmer_ci[2]), 3)
)

kable(spec_df, booktabs = TRUE,
      col.names = c("Model", "ATE", "CI lower", "CI upper"),
      caption = "Model robustness across specification choices.") %>%
  kable_styling(latex_options = "hold_position")
```

\newpage

# Heterogeneity Analysis (Conditional ATE)

## Interaction Model

```{r hte-model}
hte_form <- as.formula(
  paste("y ~ z * (race_collapsed + gender + fgen) +",
        "selfrpt + urban_f + mindset + test + sch_race + pov + size +",
        "(1 | schoolid)")
)
hte_fit <- lmer(hte_form, data = data, REML = FALSE,
                control = lmerControl(optimizer = "bobyqa"))

hte_summ   <- summary(hte_fit)$coefficients
int_terms  <- rownames(hte_summ)[grepl("^z:", rownames(hte_summ))]
int_pvals  <- hte_summ[int_terms, "Pr(>|t|)"]
int_adj    <- p.adjust(int_pvals, method = "BH")

int_table <- data.frame(
  Term        = int_terms,
  Estimate    = round(hte_summ[int_terms, "Estimate"], 3),
  SE          = round(hte_summ[int_terms, "Std. Error"], 3),
  p_raw       = round(int_pvals, 4),
  p_BH        = round(int_adj, 4),
  Significant = ifelse(int_adj < 0.05, "Yes", "No")
)

kable(int_table, booktabs = TRUE, row.names = FALSE,
      col.names = c("Interaction term", "Estimate", "SE",
                    "Raw p", "BH-adj p", "Sig. after BH"),
      caption = "Heterogeneity interaction terms from the mixed-effects model.
                 p-values adjusted with Benjamini--Hochberg FDR correction.
                 \\textit{None of the interaction terms survive BH correction at
                 $\\alpha = 0.05$} — treatment effect heterogeneity by
                 race (collapsed), gender, and first-gen status is not
                 statistically established after correction for multiple comparisons.") %>%
  kable_styling(latex_options = c("hold_position"))
```

## Subgroup ATEs with Cluster-Bootstrap CIs

```{r subgroup-ates}
# Collapse race categories with < 2 % of sample into "other" (already done via
# race_collapsed). Subgroups: race_collapsed, gender, fgen.

# Helper: cluster-bootstrap ATE within a subgroup
subgroup_boot_ci <- function(sub_data, n_boot = 200) {
  schools_sg <- unique(as.character(sub_data$schoolid))
  if (length(unique(sub_data$z)) < 2) return(c(NA, NA, NA))

  m0   <- lm(y ~ z + selfrpt + mindset + test, data = sub_data)
  est  <- coef(m0)["z"]

  boots <- vapply(seq_len(n_boot), function(i) {
    s_schools <- sample(schools_sg, length(schools_sg), replace = TRUE)
    bd <- do.call(rbind, lapply(seq_along(s_schools), function(j) {
      d <- sub_data[as.character(sub_data$schoolid) == s_schools[j], ]
      d$schoolid <- factor(j)
      d
    }))
    tryCatch(coef(lm(y ~ z + selfrpt + mindset + test, data = bd))["z"],
             error = function(e) NA_real_)
  }, numeric(1))

  ci <- quantile(boots, c(0.025, 0.975), na.rm = TRUE)
  c(ATE = unname(est), CI_lo = unname(ci[1]), CI_hi = unname(ci[2]))
}

set.seed(42)
# race_collapsed subgroups
sg_race <- data %>%
  group_by(Subgroup = paste0("Race: ", race_collapsed)) %>%
  group_modify(~ {
    r <- subgroup_boot_ci(.x)
    data.frame(N = nrow(.x), ATE = r["ATE"], CI_lo = r["CI_lo"], CI_hi = r["CI_hi"])
  }) %>% ungroup()

# gender subgroups
sg_gender <- data %>%
  group_by(Subgroup = paste0("Gender: ", gender)) %>%
  group_modify(~ {
    r <- subgroup_boot_ci(.x)
    data.frame(N = nrow(.x), ATE = r["ATE"], CI_lo = r["CI_lo"], CI_hi = r["CI_hi"])
  }) %>% ungroup()

# fgen subgroups
sg_fgen <- data %>%
  group_by(Subgroup = paste0("First-gen: ", fgen)) %>%
  group_modify(~ {
    r <- subgroup_boot_ci(.x)
    data.frame(N = nrow(.x), ATE = r["ATE"], CI_lo = r["CI_lo"], CI_hi = r["CI_hi"])
  }) %>% ungroup()

sg_all <- bind_rows(sg_race, sg_gender, sg_fgen) %>%
  filter(!is.na(ATE)) %>%
  mutate(across(c(ATE, CI_lo, CI_hi), round, 3))

# Overall ATE line for reference
overall_ate <- lmer_coef
```

```{r subgroup-forest-plot, fig.height=5.5}
ggplot(sg_all, aes(x = ATE, y = fct_rev(Subgroup))) +
  geom_point(size = 3, colour = "#22355A") +
  geom_errorbarh(aes(xmin = CI_lo, xmax = CI_hi),
                 height = 0.25, colour = "#22355A", linewidth = 0.8) +
  geom_vline(xintercept = overall_ate, linetype = "dashed",
             colour = "#EE6677", linewidth = 0.9) +
  annotate("text", x = overall_ate + 0.004, y = Inf, vjust = 1.3, hjust = 0,
           label = sprintf("Overall ATE = %.3f", overall_ate),
           colour = "#EE6677", size = 3.5) +
  labs(
    title    = "Subgroup ATEs with cluster-bootstrap 95% CIs",
    subtitle = paste("No interaction survives BH correction. CIs overlap",
                     "substantially, consistent with a relatively homogeneous effect."),
    x        = "Estimated ATE (SD units)",
    y        = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

\newpage

# Discussion and Conclusion

## Final ATE Summary Table

```{r final-summary-table}
final_table <- tribble(
  ~Method, ~Estimate, ~Naive_CI, ~Cluster_CI,

  "OLS Regression",
  round(ols_est, 3),
  make_ci_str(ols_naive_ci[1], ols_naive_ci[2]),
  make_ci_str(ci_boot(boot_ols)[1], ci_boot(boot_ols)[2]),

  "IPW (stabilised)",
  round(ipw_est, 3),
  make_ci_str(ipw_naive_ci[1], ipw_naive_ci[2]),
  make_ci_str(ci_boot(boot_ipw)[1], ci_boot(boot_ipw)[2]),

  "Doubly Robust",
  round(dr_est, 3),
  make_ci_str(dr_naive_ci[1], dr_naive_ci[2]),
  make_ci_str(ci_boot(boot_dr)[1], ci_boot(boot_dr)[2]),

  "Lasso-IPW",
  round(lasso_est, 3),
  make_ci_str(lasso_naive_ci[1], lasso_naive_ci[2]),
  make_ci_str(lasso_cl_ci[1], lasso_cl_ci[2]),

  "PSM + Regression",
  round(psm_est, 3),
  make_ci_str(psm_naive_ci[1], psm_naive_ci[2]),
  make_ci_str(psm_cl_ci[1], psm_cl_ci[2]),

  "Multilevel (lmer)",
  round(lmer_coef, 3),
  "— (inherently multilevel)",
  make_ci_str(ci_boot(boot_lmer)[1], ci_boot(boot_lmer)[2]),

  "Bayesian (MCMCregress)",
  round(post_mean, 3),
  "— (posterior)",
  make_ci_str(post_ci[1], post_ci[3])
)

kable(final_table, booktabs = TRUE,
      col.names = c("Method", "Estimate", "Naive 95% CI", "Cluster-Robust 95% CI"),
      caption = "\\textbf{Primary results table.} ATE estimates and 95\\% CIs across
                 all methods. Cluster-robust CIs account for the 76-school multilevel
                 structure. All estimates are in standard deviation units of the
                 synthetic achievement outcome. \\textit{This is a synthetic
                 dataset; the true DGP effect is approximately 0.41 SD by design.}") %>%
  kable_styling(latex_options = c("hold_position", "scale_down")) %>%
  column_spec(4, bold = TRUE) %>%
  row_spec(6, background = "#EEF4FF")   # highlight lmer as primary
```

## Methods Forest Plot

```{r final-forest-plot, fig.height=5}
forest_df <- final_table %>%
  # Parse CI strings back to numbers for the plot
  mutate(
    CI_lo = case_when(
      Method == "OLS Regression"       ~ ci_boot(boot_ols)[1],
      Method == "IPW (stabilised)"     ~ ci_boot(boot_ipw)[1],
      Method == "Doubly Robust"        ~ ci_boot(boot_dr)[1],
      Method == "Lasso-IPW"            ~ lasso_cl_ci[1],
      Method == "PSM + Regression"     ~ psm_cl_ci[1],
      Method == "Multilevel (lmer)"    ~ ci_boot(boot_lmer)[1],
      Method == "Bayesian (MCMCregress)" ~ post_ci[1]
    ),
    CI_hi = case_when(
      Method == "OLS Regression"       ~ ci_boot(boot_ols)[2],
      Method == "IPW (stabilised)"     ~ ci_boot(boot_ipw)[2],
      Method == "Doubly Robust"        ~ ci_boot(boot_dr)[2],
      Method == "Lasso-IPW"            ~ lasso_cl_ci[2],
      Method == "PSM + Regression"     ~ psm_cl_ci[2],
      Method == "Multilevel (lmer)"    ~ ci_boot(boot_lmer)[2],
      Method == "Bayesian (MCMCregress)" ~ post_ci[3]
    ),
    Primary = Method == "Multilevel (lmer)"
  )

ggplot(forest_df, aes(x = Estimate, y = fct_rev(Method),
                       colour = Primary, shape = Primary)) +
  geom_point(size = 3.5) +
  geom_errorbarh(aes(xmin = CI_lo, xmax = CI_hi),
                 height = 0.3, linewidth = 0.9) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50") +
  scale_colour_manual(values = c("FALSE" = "#4477AA", "TRUE" = "#AA3333"),
                       guide = "none") +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 18),
                      guide = "none") +
  scale_x_continuous(limits = c(0, 0.6), breaks = seq(0, 0.6, 0.1)) +
  labs(
    title    = "Forest plot: ATE estimates with cluster-robust 95% CIs",
    subtitle = "Red diamond = primary estimator (multilevel lmer). All methods use cluster-robust inference.",
    x        = "ATE (SD units of achievement outcome)",
    y        = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

## Key Findings

On the Carvalho et al. (2019) ACIC 2018 synthetic dataset derived from the National
Study of Learning Mindsets, five causal estimators converge on an ATE of approximately
**`r round(lmer_coef, 2)` SD** with cluster-robust 95% CIs of approximately
**`r make_ci_str(ci_boot(boot_lmer)[1], ci_boot(boot_lmer)[2])`**
(multilevel bootstrap; see table above for all methods).

Key methodological findings:

1. **Clustering matters.** Naive CIs that treat students as independent observations
   are meaningfully narrower than cluster-robust CIs. The school-level ICC of
   approximately `r round(icc, 2)` confirms non-trivial within-school correlation.

2. **Method agreement.** All five estimators (OLS, IPW, DR, Lasso-IPW, PSM) produce
   estimates within 0.02 SD of each other, providing strong cross-method consistency.

3. **Sensitivity.** The result is robust to unmeasured confounding up to
   Γ = `r gamma_break` (Rosenbaum bounds), and the E-value of
   `r round(ev_point, 2)` means a confounder would need to be very strongly
   associated with both treatment and outcome to explain away the effect.

4. **Heterogeneity.** After BH correction, no subgroup interaction (race-collapsed,
   gender, first-gen) survives at α = 0.05. Subgroup ATEs are consistent with a
   relatively homogeneous treatment effect across the observed groups.

5. **Honest framing.** The ~0.41 SD effect is a property of the synthetic DGP, not
   a real empirical finding. The actual NSLM trial (Yeager et al., 2019) found
   ~0.10 SD on GPA. This analysis demonstrates methodological competence on a
   known, well-studied dataset.

\newpage

# Appendix A: Bayesian MCMC Diagnostics

```{r appendix-trace, fig.height=5}
# Trace plot for z coefficient
z_chain <- as.data.frame(bayes_fit[, z_col])
names(z_chain) <- "z_coef"
z_chain$iter   <- seq_len(nrow(z_chain))

p_trace <- ggplot(z_chain, aes(x = iter, y = z_coef)) +
  geom_line(colour = "#4477AA", alpha = 0.7, linewidth = 0.3) +
  geom_hline(yintercept = post_mean, colour = "#EE6677", linetype = "dashed") +
  labs(title = "Trace plot: treatment effect coefficient (z)",
       subtitle = "Chain should look like 'fuzzy caterpillar' -- no trends or long runs",
       x = "MCMC iteration (post-burnin, thinned)", y = "Posterior draw") +
  theme_bw(base_size = 11)

# Autocorrelation plot
acf_vals <- acf(z_chain$z_coef, lag.max = 20, plot = FALSE)
acf_df   <- data.frame(lag = acf_vals$lag[-1], acf = acf_vals$acf[-1])

p_acf <- ggplot(acf_df, aes(x = lag, y = acf)) +
  geom_col(fill = "#4477AA", width = 0.5) +
  geom_hline(yintercept = c(-1.96, 1.96) / sqrt(nrow(z_chain)),
             linetype = "dashed", colour = "#EE6677") +
  labs(title = "Autocorrelation: z coefficient",
       x = "Lag", y = "ACF") +
  theme_bw(base_size = 11)

p_trace / p_acf
```

```{r appendix-geweke}
# Geweke diagnostic: compares first 10% and last 50% of chain
# |Z| < 1.96 → no convergence concern
geweke_all <- geweke.diag(bayes_fit)
geweke_df  <- data.frame(
  Parameter = names(geweke_all$z),
  Geweke_Z  = round(as.numeric(geweke_all$z), 3),
  Flag      = ifelse(abs(as.numeric(geweke_all$z)) > 1.96, "CHECK", "OK")
) %>% arrange(desc(abs(Geweke_Z)))

# Show top 10 by |Z|
kable(head(geweke_df, 10), booktabs = TRUE, row.names = FALSE,
      caption = "Geweke diagnostic Z-scores (top 10 by absolute value).
                 $|Z| < 1.96$ indicates no stationarity concern.") %>%
  kable_styling(latex_options = "hold_position")
```

```{r appendix-ess}
# Effective sample size for all parameters
ess_all <- effectiveSize(bayes_fit)
cat(sprintf("Minimum ESS (all params): %.0f\n", min(ess_all)))
cat(sprintf("z coefficient ESS:        %.0f\n", ess_z))
cat("ESS should be >> 200 for reliable inference.\n")
```

\newpage

# Appendix B: Session Information

```{r session-info}
sessionInfo()
```
